---
title: "EDA_Rhine_River"
author: "mroe1"
date: "16/04/2020"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: united
    pandoc_args: 
      ["--number-sections",
      "--number-offset=1"]
---

```{r, include = FALSE}
library(data.table)
library(ggplot2)
library(dplyr)
library(plotly)

```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE}
getwd()
```

# Exploratory data analysis

In my exploratory data analysis of the river Rhine, I produced tables and graphs on various data sets of discharge at different stations along the course of the river Rhine.
The anaysis hypothesis, suggests falling summer discharge and increasesing winter discharge, due to climate change and it knock on effects on snow depths in mountains regions and incresed evapotranspiration during summer months.

## Summary of statistical values 
brocken down by various months and various stations along the river, it is clear were data is missing due to -999 value

```{r runoff_day_raw, echo = FALSE}
runoff_stats <- readRDS("./data/runoff_stats.rds")
runoff_day <- readRDS("./data/runoff_day_raw.rds")
runoff_month <- readRDS("./data/runoff_month.rds")
runoff_stations <- readRDS("./data/runoff_stations_raw.rds")
runoff_stats$median <- runoff_day[,median(value), by = sname][,2]
runoff_stats_tidy <- melt(runoff_stats, id.vars = 'sname', by = sname)
knitr::kable(runoff_stats_tidy, caption = "summary statistics", digits = 0)

```

## Summary statistics plotted
This table shows a plots of summary statitics of all stations over the period of all data collected, it shows the spread of the data. This could be important in being able to catagorise stations into specific subsets, such as differences in mean values, or the differences in the range, hence patterens could be deduced. Data with incorrect values from previous table has been removed.

```{r barplot,echo = FALSE, fig.width = 10, fig.height = 6}
runoff_stats <- readRDS("./data/runoff_stats.rds")
runoff_day <- readRDS("./data/runoff_day_raw.rds")
runoff_month <- readRDS("./data/runoff_month.rds")
runoff_stations <- readRDS("./data/runoff_stations_raw.rds")
runoff_stats$median <- runoff_day[,median(value), by = sname][,2]
runoff_stats <- runoff_stats[sname != "REES" & sname != "MAXA" & sname != "DOMA" & sname != "FELS"]
runoff_stats_tidy <- melt(runoff_stats, id.vars = 'sname')
ggplot(runoff_stats_tidy, aes(x = sname, y = value, shape = "cyl", color = "cyl", size = 0.5)) + 
  geom_point(aes(col = variable, shape = variable)) +
   ylab(label = " Daily Runoff (m3/s)") 
```

## Each station date over whole period
To show and identify points in time when maximum points and minimum points were reached in the stations run off 
data, I created interactive plots of each station, it is also easy to see if there is correlation between station or not. It can also be seen were data is missing.

```{r, echo = FALSE, fig.width = 25, fig.height = 6}

runoff_month <- readRDS("./data/runoff_month.rds")
runoff_month <- runoff_month[value >= 0]
runoff_month[1:3654,3] <- "REES"
runoff_month <- runoff_month[sname != "<NA>"]

p <- runoff_month %>%
  ggplot( aes(x = date, y = value)) +
  geom_area(fill="#69b3a2", alpha=0.5) +
  geom_line(color="#69b3a2") +
  ylab("discharge m^3/s") +
  facet_wrap(~ sname, scales = 'free')
p <- ggplotly(p)
p
```

## Analysis of specifc stations for each month
This graph shows changes in total monthly runoff at three selected measurement stations, at different points of altitude along the river, DOMA and BASR are at higher altitudes, while KOEL is at a lower altitude, in the results it is easy to show the greater divergence of the mean at stations with higher altitude,(smaller in summer, greater in winter), this holds true for all stations. In selected different points on the river rhine, it would be important to ask such questions as, do other tributaries change overall results? And has there been any human activity near a station which would have an impact on the results ?


```{r, echo = FALSE}
runoff_month_key <- readRDS('data/runoff_month_key.rds')
runoff_year_key <- readRDS('data/runoff_year_key.rds')
colset_4 <-  c("#D35C37", "#BF9A77", "#D6C6B9", "#97B8C2")
year_thres <- 2000
runoff_year_key <- runoff_year_key[value >= 0]  
runoff_month_key <- runoff_month_key[value >= 0]  
runoff_year_key[year < year_thres, period := factor('pre_2000')]
runoff_year_key[year >= year_thres, period := factor('aft_2000')]
runoff_month_key[year < year_thres, period := factor('pre_2000')]
runoff_month_key[year >= year_thres, period := factor('aft_2000')]
ggplot(runoff_month_key, aes(factor(month), value, fill = period)) +
  geom_boxplot() +
  facet_wrap(~sname, scales = 'free_y') +
  scale_fill_manual(values = colset_4[c(4, 1)]) +
  xlab(label = "time period") +
  ylab(label = "Runoff (m3/s)") +
  theme_bw()

```

## Average run off at specific stations 
This graph shows that post 2000, the range of values of runoff are getting smaller at the sample stations ,moreover fewer outliers occur, whilst there is little significant change in the mean, this suggest the analysis hypothesis is correct, which predicted higher winter and lower summer runoff over the past 20 years relative to the whole time period of the data set

```{r, echo=FALSE}
##little change in mean values pre 2000 and post 2000
ggplot(runoff_month_key, aes(period, value, fill = period)) +
  geom_boxplot() +
  facet_wrap(~sname, scales = 'free_y') +
  scale_fill_manual(values = colset_4[c(4, 1)]) +
  xlab(label = "time period") +
  ylab(label = "Runoff (m3/s)") +
  theme_bw()
```

## Futher possible investigations
human inpacts on the river rhine have been extensive, from building of dams to use of water in waste treatment and agriculture, these influnces will have impacts on river runoff at different locations, hence identifing were and when dams and agriculutral impacts will be important in determining the run off as a function of time and displacement. Other major rivers in europe such as the elbe could also be analysed to see if it follows similar patterns as the rhine.

https://rpubs.com/mr2214/600582